 

### 2020-10-08 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

**Authors (so far, not necessarily in final order):** 
Nidhi Vinod^1^,
Martijn Slot^2^,
Ian McGregor^3^,
Elsa Ordway,
Marielle N. Smith^3^,
Tyeen Taylor,
Lawren Sack,
Kristina J. Anderson-Teixeira^1,2^*

@@

4. Department of Forestry, Michigan State University, East Lansing, MI 48824, USA

@@

Temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018;@huangAirTemperatureOptima2019]. Loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], this has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013]

@@

We're seeing increasing evidence that this impacts the ecology, with potential feedbacks to climate change, especially in protecting cool-acclimated species [@zellweger_forest_2020; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013; @defrenneMicroclimateModeratesPlant2013]

@@

**However, we lack a systematic, up to date understanding of biophysical and biological patterns across this gradient, how these affect leaf-level processes, and in turn how it affects ecology (Fig. 1).**

@@

i.	How does the biophysical environment vary with height in forests?

ii.	How do leaf traits vary with height (or between sun and shade leaves) in forests?

iii.	How do biophysical environment and traits combine to affect leaf temperature?

@@

v.	What are the implications of these patterns for the ecology and climate change responses of canopy versus understory trees? What are the implications for modeling/ scaling? (*nerge these better*)

@@

Light conditions, specifically photosynthetically active radiation (PAR) strongly varies  with canopy height due to cumulative layers of vegetation that limits light to lower canopy layers [@fauset_differences_2018; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
(Fig. **2x**)such that the forest floor of understory regions receive 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; Chazdon & Fetcher 1984; Parker et al. 2019]
In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019]. 
This difference in light regime is an important variable in determining leaf traits along the gradient and consequently their responses to temperature [@fausetDifferencesLeafThermoregulation2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008].
-	Sunflecks: Leaky et al. 2003--from Marielle, "This is another, higher resolution axis of the variability of microenvironments that is probably quite important to address. (Also see the papers by Robin Chazdon on temporal variability of light environments). Maybe we will want to add these sentences (above) and something about sunflecks to a separate section that specifically addresses sources of variation in forest structure & microenvironments (e.g., vertical, horizontal, seasonal, diurnal, and with disturbance)?"

@@

RH, in combination with temperature, determines vapor pressure deficit, $VPD$
- $VPD$ is lower in understory of Atlantic forest during wet season [@fauset_differences_2018, Tymen et al. 2017]
This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomtal contrains and lower osmotic potential, making canopy leaves more efficient in regulating water use [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]

@@

Leaf traits vary along the vertical canopy profile following the pattern of light availability across the gradient [@poorterMetaanalysisPlantResponses2019b]. Leaves in the upper canopy that are exposed to more light differ in their traits compared to shaded leaves exposed to less light [@niinemetsWorldwideAnalysisWithincanopy2015].
Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [Chen et al. 2020], leaf thermoregulation and thermal sensitivty which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018, @michaletzEnergeticCarbonEconomic2016; @niinemetsWorldwideAnalysisWithincanopy2015, Chen et al. 2020, REFS)

@@

## Scaling & implications

@@

### Representation in models

@@

>>>>

**Authors (so far, not necessarily in final order):** 
Nidhi Vinod^1^,
Martijn Slot^2^,
Ian McGregor^3^,
Elsa M. Ordway^4^,
Marielle N. Smith^5^,
Tyeen Taylor,
Lawren Sack,
Kristina J. Anderson-Teixeira^1,2^*

@@

4. Department of Organismic and Evolutionary Biology, Harvard University, Cambridge, MA 02138, USA

5. Department of Forestry, Michigan State University, East Lansing, MI 48824, USA

@@

Temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018;@huangAirTemperatureOptima2019]. Loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], this has been shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013]

@@

We're seeing increasing evidence that this impacts ecosystems, with potential feedbacks to climate change, especially in protecting species acclimated to cooler [mean annual] temperatures [@zellweger_forest_2020; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013; @defrenneMicroclimateModeratesPlant2013]

@@

**However, we lack a systematic, up to date understanding of biophysical and biological patterns across this gradient, how these affect leaf-level processes, and in turn how it affects ecosystems (Fig. 1).**

@@

i.	How does the biophysical environment vary with height in forests? ("What aspects of biophysical envt? Microclimate and vertical plant structure/phenology/etc (ok yes, see this in fig 1)"?--Elsa:
So change to micrometeorological environment? Not biological just physical in this question, right?

ii.	How do leaf traits vary with height (or between sun and shade leaves) in forests?(What types of leaf traits?-E.O)

iii. How do vertical changes in biophysical environment conditions and leaf traits combine to affect leaf temperature [and in turn photosynthetic capacity]??

@@

v. What are the implications of these patterns for the ecology ecosystem and climate change responses of canopy versus understory trees?

vi.	What are the implications for modeling/ scaling / What are the implications of these leaf-scale responses at the ecosystem scale [in the context of climate change] / How do these leaf-level responses scale to entire ecosystems(*nerge these better*)

@@

Light conditions, specifically photosynthetically active radiation (PAR) strongly varies  with canopy height due to cumulative layers of vegetation that limits light to lower canopy layers [@fauset_differences_2018; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
(Fig. **2x**), such that the forest floor of understory regions receive 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; Chazdon & Fetcher 1984; Parker et al. 2019].
In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019]. 
This difference in light regime is an important variable in determining leaf traits along the gradient and consequently their responses to temperature [@fausetDifferencesLeafThermoregulation2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008].
-	Sunflecks: Leaky et al. 2003--from Marielle, "This is another, higher resolution axis of the variability of microenvironments that is probably quite important to address. (Also see the papers by Robin Chazdon on temporal variability of light environments). Maybe we will want to add these sentences (above) and something about sunflecks to a separate section that specifically addresses sources of variation in forest structure & microenvironments (e.g., vertical, horizontal, seasonal, diurnal, and with disturbance)?"

@@

RH, in combination with temperature, determines vapor pressure deficit, $VPD$
- $VPD$ is lower in understory of Atlantic forest during wet season [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017]
This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomtal contrains and lower osmotic potential, making canopy leaves more efficient in regulating water use [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]

@@

Leaf traits vary along the vertical canopy profile following the pattern of light availability across the gradient [@poorterMetaanalysisPlantResponses2019b]. Leaves in the upper canopy that are exposed to more light differ in their traits compared to shaded leaves exposed to less light [@niinemetsWorldwideAnalysisWithincanopy2015].
Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [Chen et al. 2020], leaf thermoregulation and thermal sensitivty which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016; @niinemetsWorldwideAnalysisWithincanopy2015, @chenLeafEconomicsSpectrum2020, REFS)

@@

## Ecosystem Scaling & implications

@@

## Implications for Modeling / Scaling / Future Climate Projections

Elsa--The content below could probably be integrated in the above [Ecosystem Scale Implications] section, with subsections focusing on: 
-	Suggested implications at the ecosystem scale based on literature
-	How do models currently integrate or not integrate these concepts and dynamics
-	What major gaps in knowledge and/or important questions emerge from this review (i.e. a lot of focus on droughts justifiably, but understanding heterogeneous impacts of temperature increases on forest ecosystems, particularly at vertically resolved scales is also important.)
-	What does that mean for about ability to predict/understand how forest ecosystems will respond in the future?

@@

--Variation across horizontal biophysical gradients (climate, soils, etc.)(E.O)
